---
title: "STOmics Stereo-seq introduction"
author: John Salamon
editor_options:
  chunk_output_type: console
---

# Stereo-seq background

Stereo-seq (SpaTial Enhanced Resolution Omics-sequencing) builds upon the chip technology already used in MGI DNA nanoball sequencing (DNBSEQ). For background on DNBSEQ, see [here](https://mgi-tech.eu/technology).

More information on Stereo-seq can be found in the [paper]("https://www.cell.com/cell/fulltext/S0092-8674(22)00399-3")

TODO: would be nice to include images or maybe just link to some slides

# Data processing basics

Once sequencing is complete and you have fastq files, you can perform basic analysis using the Stereo-seq Analysis Workflow (SAW).
This command-line tool is more or less the STOmics equivalent of Spaceranger.

As of time of writing, the latest version of SAW is [v8.1 (see documentation)](https://stereotoolss-organization.gitbook.io/saw-user-manual-v8.1).

In addition, `saw --help` and subcommands (e.g. `saw count --help`) provide helpful clues for usage.

## Obtaining SAW

You can download the latest version of SAW from the [STOmics download page](https://en.stomics.tech/products/stomics-software/stomics-offline-software/list.html) (does require making an account).

Alternatively, you can find earlier versions of SAW on [Github](https://github.com/STOmics/SAW).
SAW has changed significantly between versions 7.x and 8.x. 
Version 8.x introduces both internal workflow management with Snakemake, and compatibility with OMNI (FFPE-compatible Stereo-seq).


## SAW makeRef

The first step of your analysis will be to choose and [build an appropriate reference](https://stereotoolss-organization.gitbook.io/saw-user-manual-v8.1/tutorials/preparation-of-reference). This step can mostly be completed with `saw makeRef`.

To build the reference for this example, here's the command I used:

```{bash, eval=FALSE}
#!/bin/bash

saw makeRef \
    --mode=STAR \
    --fasta=/path/to/GRCm39.genome.fa \
    --gtf=/path/to/annotation.gtf \
    --genome=/path/to/reference/output
```

Notes:

- The [GENCODE GRCm39 reference](https://www.gencodegenes.org/mouse/) was used
- The `--genome` parameter actually specifies the name of the output directory
- There is no way to pass parameters to STAR in this version.

To have more control over the parameters used, you can use version 7.x of SAW [from Github](https://github.com/STOmics/SAW). This version is implemented as a bash
script which calls into a container image, and is easier to customise.


## SAW count


Next, we want to run the main pipeline which is called via `saw count`.

Documentation for this step is available for using SAW with different sample types, e.g. [fresh-frozen](https://stereotoolss-organization.gitbook.io/saw-user-manual-v8.1/tutorials/run-main-pipeline/stereo-seq-ff).

Here's how it was run for this demo:

```{bash, eval=FALSE}
#!/bin/bash

saw count \
    --id=workshop_demo \
    --sn=C04042E3 \
    --chip-mask="/path/to/C04042E3.barcodeToPos.h5" \
    --omics=transcriptomics \
    --kit-version="Stereo-seq T FF V1.3" \
    --sequencing-type="PE75_50+100" \
    --organism=mouse \
    --tissue=brain \
    --fastqs=/path/to/fastq/directory/ \
    --image-tar="/path/to/C04042E3_SC_20240925_142342_4.1.0.tar.gz" \
    --reference=/path/to/reference/output
```

Notes:

- `--reference` here should point to the directory specified in the `--genome` parameter of `saw makeRef`
- `--image-tar` accepts image files processed by [StereoMap](#stereomap) or ImageStudio, and is optional
- `--fastqs` points to a directory containing fastq file pairs. More than one pair can be present (e.g. if multiple sequencing runs or lanes were used) 
- `--organism` and `--tissue` can be any string

# SAW outputs

Once this is complete, an `outs` directory will be created:

```{bash, eval=FALSE}
$ tree outs/
.
├── analysis
│   ├── C04042E3.bin20_1.0.h5ad
│   ├── C04042E3.bin20_1.0.marker_features.csv
│   ├── C04042E3.bin50_1.0.h5ad
│   ├── C04042E3.bin50_1.0.marker_features.csv
│   ├── C04042E3.cellbin_1.0.adjusted.h5ad
│   ├── C04042E3.cellbin_1.0.adjusted.marker_features.csv
│   ├── C04042E3.cellbin_1.0.h5ad
│   └── C04042E3.cellbin_1.0.marker_features.csv
├── bam
│   └── annotated_bam
│       ├── C04042E3.Aligned.sortedByCoord.out.merge.q10.dedup.target.bam
│       └── C04042E3.Aligned.sortedByCoord.out.merge.q10.dedup.target.bam.csi
├── C04042E3.report.tar.gz
├── feature_expression
│   ├── C04042E3.adjusted.cellbin.gef
│   ├── C04042E3.cellbin.gef
│   ├── C04042E3.merge.barcodeReadsCount.txt
│   ├── C04042E3_raw_barcode_gene_exp.txt
│   ├── C04042E3.raw.gef
│   └── C04042E3.tissue.gef
├── image
│   ├── C04042E3_HE_mask_edm_dis_10.tif
│   ├── C04042E3_HE_mask.tif
│   ├── C04042E3_HE_regist.tif
│   └── C04042E3_HE_tissue_cut.tif
├── visualization.tar.gz
└── visualization.tar.gz.md5
```

[See here](https://stereotoolss-organization.gitbook.io/saw-user-manual-v8.1/analysis/outputs/count-outputs) for a detailed guide to these outputs.

Many of these are actually HDF5 files (including the `.h5ad` (anndata-compatible analysis results) and `.gef` (count matrices) files) and can be inspected with any HDF5 compatible tool or library.

## HTML report

The first place to look is generally the HTML report. A detailed guide to the SAW report can be found [in the SAW manual](https://stereotoolss-organization.gitbook.io/saw-user-manual-v8.1/analysis/outputs/html-report).


## visualization.tar.gz

This packages the main outputs into a single file for visualization.
It is also a handy file for distributing results easily, as it packages most
of the main outputs (excluding the BAM files).

Taking a look inside, we see many of the other output files have been repackaged here:

```{bash, eval=FALSE}
.
├── C04042E3.adjusted.cellbin.gef
├── C04042E3.bin20_1.0.h5ad
├── C04042E3.bin50_1.0.h5ad
├── C04042E3.cellbin_1.0.adjusted.h5ad
├── C04042E3.rpi
├── C04042E3_SC_20240925_142342_4.1.0.tar.gz
├── C04042E3.stereo
├── C04042E3.tissue.gef
└── HE_matrix_template.txt
```

A JSON manifest of the experiment can be found in `.stereo`, which allows loading
all of the packaged data in a single step from StereoMap.

# StereoMap

StereoMap is an interactive tool to examine your Stereo-seq data (similar to Loupe Browser from 10x).

StereoMap can be [downloaded from STOmics directly](https://en.stomics.tech/products/stomics-software/stomics-offline-software/list.html), although it is currently only compatible with Windows.

StereoMap v4.1 is compatible with SAW v8.1. Previous versions also included
image QC in a seperate program, ImageStudio, which is now integrated with 
StereoMap.

# StereoPy

[StereoPy](https://stereopy.readthedocs.io/en/latest/content/01_Usage_principles.html) is a Python library designed to work with Stereo-seq data. Internally it wraps libraries such as [scanpy](https://scanpy.readthedocs.io/en/stable/), [squidpy](https://squidpy.readthedocs.io/en/stable/) and [anndata](https://anndata.readthedocs.io/en/stable/).

As most SAW outputs are anndata compatible, it is very easy to import Stereo-seq data into the Python ecosystem.

TODO: show example loading code

# Seurat

The easiest way to load Stereo-seq data in Seurat is to use the [conversion script included with StereoPy](https://stereopy.readthedocs.io/en/latest/Tutorials/Format_Conversion.html#Working-with-Seurat).

# SpatialExperiment

[SpatialExperiment](https://www.bioconductor.org/packages/release/bioc/vignettes/SpatialExperiment/inst/doc/SpatialExperiment.html) library is an S4 class that extends SingleCellExperiment. 

We can create this object manually with the `rhdf5` library.


```{r}
# for extracting data from H5 archive
library(rhdf5)
library(Matrix)
library(S4Vectors)

# spatial libraries
library(SpatialExperiment)
library(ggspavis)
```


## Loading the matrix

First we need to read from the `.h5ad` file.

```{r}

# name our input file
h5ad_file <- "data/C04042E3.bin50_1.0.h5ad"

# Read in the matrix data (CSR format)
matrix_data <- h5read(h5ad_file, "X")

# Extract dimensions (this is stored as an attribute)
shape <- h5readAttributes(h5ad_file, "X")$shape

```

Next, we convert the matrix data into a format compatible with R.

```{r}

# Extract the components
data <- matrix_data$data        # Non-zero values
indices <- matrix_data$indices  # Column indices of non-zero values
indptr <- matrix_data$indptr    # Row pointers for non-zero values

# Reconstruct the CSR matrix (have to add 1 to indices for R)
csr_matrix <- sparseMatrix(
  i = as.integer(rep(seq_len(length(indptr)-1), diff(indptr))),  # expand row indices
  j = as.integer(indices + 1),  # R uses 1-based indexing, so add 1 to column indices
  x = as.numeric(data),
  dims = shape,
)
```

We can also add in row and col data.

```{r}
obs <- h5read(h5ad_file, "obs")
var <- h5read(h5ad_file, "var")
```

Now constructing a SpatialExperiment object is just a matter of passing in the required objects.
```{r}
# create SpatialExperiment object
spe <- SpatialExperiment(
  assays = list(counts = t(csr_matrix)), 
  colData = DataFrame(cluster=as.factor(obs$leiden$codes), in_tissue=TRUE), # get cluster in there 
  rowData = DataFrame(gene_id=var$`_index`),
  spatialCoords = data.matrix(DataFrame(x=obs$x, y=obs$y))
)
spe
```

We can now continue to use this object as we would for other array-based experiments.

```{r}
library(ggspavis)

# Get norm counts 
colData(spe)$sums <- colSums(counts(spe))
plotSpots(spe, annotate="sums")
```

