
```{r}
# For RCTD
library(spacexr)

# spatial experiment
library(SpatialExperiment)

# Easier plotting with spe
library(ggspavis)

# Custom functions
source("code/stomicsFunctions.R")
```

# LISTA data (DY1_D0_stereo-seq.h5ad)

## Spatial area

DY1_D0 represents $48.143 mm^2$

From the paper: each bin contains $50x50$ nanoballs.

Center-to-center distance of $715 nm$ between nanoballs, aka bin-resolution of $36x36 um$.

Liver cells are larger, so this is probably capturing close to a single-cell diameter.

## Experimental design

- In total, 30 whole liver lobe sections
- Time series (Day 0,1,2,3,7)
- Matched scRNA-seq.

This experiment has matched and annotated single cell data. We'll be using 
this to generate Figure 1A from the paper using robust cell-type decomposition
(RCTD).

## Loading the LISTA dataset 

As they are stored as AnnData, we can import the files from this experiment in 
the same way we would files directly output from SAW.

```{r}
# SPE
h5ad_file <- "stomics/raw/DY1_D0_stereo-seq.h5ad"

# Read in the matrix data (CSR format)
matrix_data <- h5adMatrixLoad(h5ad_file)

spatial_coords <- h5read(h5ad_file, "/obsm/spatial")

rowdata <- anndataDataframe(h5ad_file, "var")
coldata <- anndataDataframe(h5ad_file, "obs")


spatial_coords <- data.matrix(data.frame(
  x=spatial_coords[1,],
  y=spatial_coords[2,],
  row.names = rownames(coldata)
))

spe <- SpatialExperiment(
  assays = list(counts = matrix_data),
  colData = coldata, # spatial bins
  rowData = rowdata, # features
  spatialCoords = spatial_coords
)

spe
```
We can quickly plot this to verify it looks correct using `plot`.
`ggplot` is much more powerful but sometimes we just want to verify the data
is there and more or less looks correct.

```{r}
# R's default plot function is actually pretty smart
# but ggplot gives more functionality
plot(spatialCoords(spe))
```
We can flip the y axis and set the aspect ratio correctly, for a start.

```{r}
# flip y axis, and add colours
ggplot(spatialCoords(spe), aes(x, y)) +
  geom_point(size = 0.5) +
  theme(aspect.ratio=1) +
  scale_y_reverse()
```
## Examing the annotation



```{r}
# Plot the current annotation
# These were created using AddModuleScore in Seurat with lists of genes
# https://github.com/haoshijie13/LISTA/blob/main/00.cut_zonation_layer_and_pathway_module_score.R

# ggspavis plotSpots can be useful
# plotSpots(spe, annotate = "annotation", in_tissue = NULL) +
  
ggplot(spatialCoords(spe), aes(x, y, color = spe$annotation)) +
  geom_point() +
  scale_color_viridis_d() +
  theme(aspect.ratio=1) +
  scale_y_reverse()

```
To get a more accurate picture we can use `hexbin`:

```{r}
ggplot(spatialCoords(spe), aes(x, y)) +
  geom_hex(bins=80, mapping = aes(fill = spe$annotation)) +
  scale_fill_viridis_d() +
  theme(aspect.ratio=1) +
  scale_y_reverse()

```

## Loading existing single cell reference 

```{r}

# examine the file
# h5ls(sc_reference)
# str(h5read(sc_reference, "var"))


sc_reference <- "stomics/raw/Homeostasis_hepatic_cell_scRNAseq.h5ad"

# Load the raw data
sc_matrix_data <- h5adMatrixLoad(sc_reference, group = "/raw/X")
obs <- anndataDataframe(sc_reference, "obs")
var <-anndataDataframe(sc_reference, "var")

# make sure col and rownames are set
rownames(sc_matrix_data) <- var$gene
colnames(sc_matrix_data) <- rownames(obs)

sce <- SingleCellExperiment(
  assays = list(counts = sc_matrix_data),
  colData = obs, # spatial bins
  rowData = var # features
)

```

We can see how many cells we have per annotation:

```{r}
table(sce$annotation)
```



## RCTD

Robust cell type deconmposition ([RCTD](https://www.nature.com/articles/s41587-021-00830-w))


Because this will take quite some time with our full dataset, we'll actually
just take a random subsample of cells for this example.

```{r}

# Make sure we get a usable seed...
set.seed(123)

# We can subsample our sce for speed
sce <- sce[ ,sample(dim(sce)[2], 8000)]

# Extract single cell annotations
cell_types <- sce$annotation
names(cell_types) <- colnames(sce)

# Create RCTD object
reference <- Reference(assays(sce)$counts, cell_types)

```


```{r}
table(sce$annotation)
```


Now we need to prepare the spatial data.
First, let's take another subset (just for speed).

```{r}
# Let's select a little circle from the middle of the tissue
midpoint <- as.integer(colMeans(spatialCoords(spe)))

# center with midpoint around origin
adjusted_coords <- sweep(spatialCoords(spe), 2, midpoint, "-")

# define some regions
spe$circle <- sqrt(rowSums((adjusted_coords)^2)) < 40 # A circle, 50 pts in diameter
spe$square <- (rowSums(abs(adjusted_coords) < 50)) == 2 # A square

# We can plot more easily if we have x and y in coldata..
spe$x_pos <- spatialCoords(spe)[,1]
spe$y_pos <- spatialCoords(spe)[,2]

# Take the circle
spe_circle <- spe[, spe$circle]

# Plot...
ggplot(colData(spe_circle), aes(x_pos, y_pos)) +
  geom_hex(bins=30, mapping = aes(fill = annotation)) +
  scale_fill_viridis_d() +
  theme(aspect.ratio=1) +
  scale_y_reverse()
```


```{r}
coords <- spatialCoords(spe_circle)
rownames(coords) <- colnames(spe_circle)
puck <- SpatialRNA(data.frame(coords), assays(spe_circle)$counts)
```


Now, let's try running RCTD. 


```{r}
# Prepare for RCTD analysis
# The max_cores thing seems broken, it just uses total n cores regardless
RCTD <- create.RCTD(puck, reference, max_cores = 1)

# Multi mode allows us to define >2 cell types per tissue
RCTD <- run.RCTD(RCTD, doublet_mode = 'multi')

# notes: started at 10:30, finished at 10:46...

# Save the output as it can take a while to run
saveRDS(RCTD, "stomics/data/RCTD.rds")
```

We'll load our RDS file in case we weren't able to run this (or we don't want to wait):

```{r}
RCTD <- readRDS("stomics/data/RCTD.rds")

```



```{r}
# We get back a list of results
length(RCTD@results)

# But not all spots have results
dim(RCTD@spatialRNA@counts)

# Take only the top weighted result
celltypes <- lapply(RCTD@results, \(x) x$cell_type_list[which.max(x$sub_weights)])
celltypes <- unlist(celltypes)


# Empty vec, with names, correct length
vec <- rep("None", dim(spe_circle)[2])
names(vec) <- colnames(spe_circle)

# Our results
celltypes
names(celltypes) <- colnames(RCTD@spatialRNA@counts)

# assign
vec[names(celltypes)] <- celltypes

table(vec)
```
Each spot will have a list of identified cell types.
It'll also have a list of identified cells.


```{r}
results <- RCTD@results
# normalize the cell type proportions to sum to 1.
cell_type_names <- RCTD@cell_type_info$info[[2]] #list of cell type names
cell_type_names
```
```{r}

spe_circle$celltypes <- vec

# Plot...
ggplot(colData(spe_circle), aes(x_pos, y_pos, color=celltypes)) +
  geom_point() +
  scale_color_discrete() +
  theme(aspect.ratio=1) +
  scale_y_reverse()
```

In the remaining time, perhaps we could do:

- Clustering with these results? (PCA/UMAP/kmeans)
- Define some more spatially variable genes?
