---
title: "10x Xenium Analysis Part 1"
output: html_document
date: "2024-11-18"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This part of the analysis will read in raw data and construct a data object containing 10x Xenium data. It will also explore the Xenium outputs explore the data object

# Load libraries

```{r, echo=FALSE, include=FALSE}
library(SpatialFeatureExperiment)
library(Voyager)
library(RBioFormats)
library(scater)
library(BiocParallel)
library(scran)
library(fs)
library(bluster)
library(sf)
library(EBImage)
library(terra)
library(scales)
library(plotly)
```

# Read in Xenium data

* Locate the Xenium output bundle on your computer and set it to ```dir```
* Might have to unzip first
* Check the Xenium output bundle in a tree-like format

```{r}
# Xenium_V1_Human_Colon_Cancer_P5_CRC_Add_on_FFPE_outs.zip

# read data into an SE object
# dir should point to the Xenium outputs
# this output folder contains data from patient 5
dir <- "xenium/raw/"

# examine files in Xenium output bundle
fs::dir_tree(dir)

# the main files needed for construction of SE are:
#cell_feature_matrix.h5
#cells.parquet
#experiment.xenium

# create data object and call it se
se <- SpatialFeatureExperiment::readXenium(data_dir = dir, 
                                                    sample_id = "p5",
                                           row.names = "symbol")

# Read in data with readRDS instead
# se <- readRDS(~/data/se_xenium.rds)
```

# Exploring the Xenium SE object

* Getting, setting, and plotting fields of SE objects
* Common operations on SE objects
* getters, setters, and plotting functions
* Find the gene count matrix

```{r}
# summary
se
# dimensions
dim(se)

# where are the counts?
# use the 'counts' getter function 
m <- counts(se)
head(m)
```

# Explore metadata within the Xenium SE data object

* The gene count matrix has metadata about the cells and genes.
* This metadata is stored within various slots of the SE object.
* Use the ```colData``` getter function to get cell metadata and ```rowData``` to get gene metadata.

```{r}
colData(se) # cell metadata
rowData(se) # gene metadata

colGeometries(se)

# use the $ operator to access slops within the object
# se$

# save file as SE object
# re-use this later as needed
#saveRDS(object = se, file = "/DIR/se_p5.rds")
```

# Plot Xenium cells

* Extract cell segmentation results to a data frame

```{r}
(cells <- colGeometry(se, "cellSeg"))
```

* Plot cell geometries using the ```colGeometry``` function

```{r}
ggplot(cells) + geom_sf() + ggtitle("Xenium cells - p5")
```

# Plot Xenium cells custom

* Plot the cell segmentation results. Also plot with the dark theme. 
* Use the ```plotGeometry``` function. 

```{r}
# "type" or "colGeometryName" depending on version of the package
plotGeometry(se, colGeometryName = "cellSeg", show_axes = TRUE) +
    plotGeometry(se, colGeometryName = "cellSeg", show_axes = TRUE, dark = TRUE)
```

# Save data checkpoint

```{r}
#saveRDS(object = se, file = "xenium/data/se_xenium.rds")
```
